---
title: "Introduction to R"
author: "Nguyen Bich Ngoc"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  pdf_document:
    number_sections: true
papersize: a4
---

# Introduction
## R
* Developed from S - interpreted language
* mostly use in statistics and data analysis
* free & open-source
* extendable with packages and new functions
* [https://cran.r-project.org/](https://cran.r-project.org/)

## RStudio
* Interface
* only work with R installed
* free & open-source
* start and save scripts
* Ctrl+Enter to run
* [http://rstudio.com/](http://rstudio.com/)





## Let's start with some simple calculations

```{r}

2+3

4^2
sqrt(25)
8^(1/3)

pi
exp(1)

log(exp(2))
log10(100)

```


## Packages & Functions

* packages add additional functions
* packages need to be installed once 
* sometime will need to be installed again after update
* load before every run

```{r}

# install.packages("ggplot2")
library(ggplot2)


```

* functions receive arguments => return results

```{r}

rnorm(100, mean = 0, sd = 1)

hist(rnorm(100, mean = 0, sd = 1))

```

## Operators

```{r}

x <- 2+3
y = 6-5

x == 3
y != 2
x < 0
y < 4
x >=5
y <= 10
is.na(x)

x < 0 & y < 4

x < 0 | y < 4

```

## Getting help

```{r}
?sqrt
# x + 3
x <- 2
x + 3

```

# Data
## Data types

* numeric: e.g. `1`, `45.3`
* integer: e.g. `2L`, `53L`
* logical: e.g. `TRUE`, `T`, `FALSE`, `F`
* character: e.g. `"orange"`, `"female"`, `"Totally agree"`


## Data structure

* Vectors
* Matrix
* Data frame
* List

### Vectors
* simplest type

```{r}
x <- c(1, 8, 23, -7, 13)
x
y <- c("a", "b", "c", "d", "e")
y

```


* same type of data

```{r}

a <- c(1, "a", 3, T)
a 
str(a)
str(x)
str(y)

b = c(1L, 8L, 23L, -7L, 13L)
str(b)

```

* creating vector 

```{r}

x <- c(1, 8, 23, -7, 13)
x <- 1:20
y <- seq(from = 3, to = 8, by = 0.2)


rep("Female", 10)

```

* Vectorization

```{r}

x <- c(3, 7, 6, 3, 5, 2)
x + 1
x*2
sqrt(x)

```


* logical vector


```{r}
x
z <- x > 4
z
y <- c("a", "b", "a", "d", "e")

t <- y == "A"
t
t <- y == "a"
t

```

* useful functions for vectors

```{r}

x
length(x)
sum(x)
max(x)
min(x)
sort(x)
order(x)
unique(x)

mean(x)
sd(x)


y
length(y)
unique(y)

```




* subset vector

```{r}
x
x[1]
x[3:5]
x[c(1,3:5)]

x[-2]
x[x > 4]


v <- 3
v[1]

```

### Matrices

* same type of data
* columns & rows


```{r}

x <- rbind(c(1:4), c(5:8))
x
dim(x)
dimnames(x)
attributes(x)

x <- cbind(c(1:4), c(5:8))
attributes(x)

x <- matrix(1:8, nrow = 2, ncol = 4, byrow = T)
x
x <- matrix(1:8, nrow = 2, ncol = 4, byrow = F)
x

```

### Data frame

* extension of matrix
* columns of different data types


```{r}
id <- 1:6
gender <- rep(c("F", "M"), each = 3)
age <- rep(8, 6)

students <- cbind(id, gender, age)
students
str(students)
summary(students)

students <- cbind.data.frame(id, gender, age)
students
str(students)
summary(students)

id <- 1:6
math <- c(9, 6, 7, 8, 10, 5)
english <- c(8, 5, 9, 8, 9, 7)

results <- cbind.data.frame(id, math, english)

df <- merge(students, results)
df

```
* subset data frame

```{r}

sub1 <- df[1:3, 1:2]
sub1

sub2 <- df[, c(1, 4:5)]
sub2

sub3 <- df$gender
sub3

sub4 <- subset(df, math > 7, select = c(id, english))
sub4

sub5 <- subset(df, english <= 7, select = -age)
sub5

```
* reorder the data frame

```{r}

df1 <- df[order(df$math),]
df1 
```
### List

* contain different type of data
* can contain data frame
```{r}
cls1 <- data.frame(id = 1:5, names = c("Lan", "Hung", "Tuan", "Mai", "Long"))
cls2 <- data.frame(id = 6:10, names = c("Thanh", "Son", "Nghia", "Hanh", "Thuy"))

ls <- list(cls1 = cls1, cls2 = cls2)
ls

```

* subset list

``` {r}
ls[1]
ls[[1]]

ls$cls1

```


## Import data

* Direct typing
* From clipboard

```{r}
df <- read.delim("clipboard")


```

* From csv 

``` {r}

df <- read.csv("C:/Users/nbngo/OneDrive/Work/[C] Quantitative research methods/quant_rm/Data - Quant Methods/Dataset_environmental_sustainability.csv", sep = ",", header = T)

```

* From xlsx
``` {r}

# install.packages("readxl")
library(readxl)

df <- read_excel("C:/Users/nbngo/OneDrive/Work/[C] Quantitative research methods/quant_rm/Data - Quant Methods/Dataset_environmental_sustainability.xlsx")

```

* From Rdata
``` {r}

load("C:/Users/nbngo/OneDrive/Work/[C] Quantitative research methods/quant_rm/Data - Quant Methods/ntl_joined_avg.Rdata")


```

## Export data

* as Rdata
``` {r}

save(df, file = "C:/Users/nbngo/OneDrive/Work/[C] Quantitative research methods/quant_rm/Data - Quant Methods/environment_survey.Rdata")


```

* to clipboard
```{r}
write.table(ntl_joined_avg, "clipboard", sep = "\t", row.names = F)
```

* to csv
```{r}
write.csv(ntl_joined_avg, file = "C:/Users/nbngo/OneDrive/Work/[C] Quantitative research methods/quant_rm/Data - Quant Methods/ice_cover_lake.csv", row.names = F)
```


# Exploring and plotting data
## Data
* mtcars data in R


```{r}

data <- mtcars

head(mtcars)

```
* [explaining variables](https://rstudio-pubs-static.s3.amazonaws.com/61800_faea93548c6b49cc91cd0c5ef5059894.html)

* summary() gives overall information of the data

```{r}
summary(data)
```

## Numerical/continuous (ratio/interval) variables
```{r}

summary(data$mpg)

mean(data$mpg)
median(data$mpg)

sd(data$mpg)
var(data$mpg)
quantile(data$mpg, seq(0, 1, 0.2))

hist(data$mpg)

ggplot(data, aes(x = mpg)) +
  geom_histogram()

ggplot(data, aes(x = mpg)) +
  geom_histogram(binwidth = 2)

ggplot(data, aes(x = mpg)) +
  geom_histogram(binwidth = 2) +
  theme_bw()

??theme_bw

ggplot(data, aes(x = wt)) +
  geom_histogram(binwidth = 0.5) +
  theme_bw()

```
## Character/Factor (Nominal/Order/Categorical)

```{r}
table(data$cyl)
prop.table(table(data$cyl))

table(data$am)
prop.table(table(data$am))


ggplot(data, aes(x = am)) +
  geom_bar() +
  theme_bw()

data$am_f <- factor(data$am)


ggplot(data, aes(x = am_f)) +
  geom_bar() +
  theme_bw()

data$am_f <- factor(data$am, labels = c("auto", "manual"))


ggplot(data, aes(x = am_f)) +
  geom_bar() +
  theme_bw()


```

## Bivariate 
* continuous & discrete

```{r}

ggplot(data, aes(x = am_f, y = mpg)) +
  geom_bar(stat = "summary", fun = "mean") +
  theme_bw()

ggplot(data, aes(x = am_f, y = mpg)) +
  geom_bar(stat = "summary", fun = "mean", fill = "lightblue") +
  theme_bw()

ggplot(data, aes(x = am_f, y = mpg)) +
  geom_boxplot() +
  theme_bw()


# install.packages("dplyr")
library("dplyr")

data %>%
  group_by(am_f) %>%
  summarize(min = min(mpg),
            sd = sd(mpg),
            min = min(mpg),
            q1 = quantile(mpg, 0.25),
            median = median(mpg),
            q3 = quantile(mpg, 0.75),
            max = max(mpg))


```

* Continuous & continuous

```{r}

ggplot(data, aes(x = wt, y = mpg)) +
  geom_point() +
  theme_bw()


ggplot(data, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm") +
  theme_bw()


ggplot(data, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", col = "red", se = F) +
  theme_bw()


cor(data$wt, data$mpg)
cor.test(data$wt, data$mpg)

```
* Discrete & discrete

```{r}

table(data$cyl, data$am_f)


ggplot(data, aes(x = am_f, fill = cyl, group = cyl)) +
  geom_bar() +
  theme_bw()


data$cyl_f <- factor(data$cyl)

ggplot(data, aes(x = am_f, fill = cyl_f, group = cyl)) +
  geom_bar() +
  theme_bw()

ggplot(data, aes(x = am_f, fill = cyl_f, group = cyl)) +
  geom_bar(position = "dodge") +
  theme_bw()

```

## Add dimension


```{r}

ggplot(data, aes(x = wt, y = mpg, col = cyl_f)) +
  geom_point()  +
  theme_bw()

ggplot(data, aes(x = wt, y = mpg, col = cyl_f)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)  +
  theme_bw()

ggplot(data, aes(x = wt, y = mpg, col = cyl_f)) +
  geom_point() +
  facet_grid(.~am_f) +
  theme_bw()

```


# Linear regression

## Compare means


```{r}

fit1 <- lm(mpg ~ am_f, data)
summary(fit1)

fit2 <- lm(mpg ~ cyl_f, data)
summary(fit2)

```

## Simple linear regression


```{r}

fit3 <- lm(mpg ~ wt, data)
summary(fit3)

```

## Multiple linear regression


```{r}

fit4 <- lm(mpg ~ wt + cyl_f + am_f, data)
summary(fit4)

fit5 <- lm(mpg ~ wt + cyl_f , data)
summary(fit5)

AIC(fit4)
AIC(fit5)

anova(fit4, fit5)
anova(fit4, fit3)

plot(fit5)

```
