---
title: "Introduction to R"
author: "Nguyen Bich Ngoc"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: true
  html_document:
    df_print: paged
papersize: a4
---


```{r, echo=F}

library(formatR)
library(knitr)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

# Introduction
## R
* Developed from S - interpreted language
* mostly use in statistics and data analysis
* free & open-source
* extendable with packages and new functions
* [https://cran.r-project.org/](https://cran.r-project.org/)

## RStudio
* Interface
* only work with R installed
* free & open-source
* start and save scripts
* Ctrl+Enter to run
* [http://rstudio.com/](http://rstudio.com/)





## Let's start with some simple calculations

```{r}

2+3

4^2
sqrt(25)
8^(1/3)

pi
exp(1)

log(exp(2))
log10(100)

```


## Packages & Functions

* packages add additional functions
* packages need to be installed once 
* sometime will need to be installed again after update
* load before every run

```{r}

# install.packages("ggplot2")
library(ggplot2)


```

* functions receive arguments => return results

```{r}

rnorm(100, mean = 0, sd = 1)

hist(rnorm(100, mean = 0, sd = 1))

```

## Operators

```{r}

x <- 2+3
y = 6-5

x == 3
y != 2
x < 0
y < 4
x >=5
y <= 10
is.na(x)

x < 0 & y < 4

x < 0 | y < 4

```

## Getting help

```{r}
?sqrt
# x + 3
x <- 2
x + 3

```

# Data
## Data types

* numeric: e.g. `1`, `45.3`
* integer: e.g. `2L`, `53L`
* logical: e.g. `TRUE`, `T`, `FALSE`, `F`
* character: e.g. `"orange"`, `"female"`, `"Totally agree"`


## Data structure

* Vectors
* Matrix
* Data frame
* List

### Vectors
* simplest type

```{r}
x <- c(1, 8, 23, -7, 13)
x
y <- c("a", "b", "c", "d", "e")
y

```


* same type of data

```{r}

a <- c(1, "a", 3, T)
a 
str(a)
str(x)
str(y)

b = c(1L, 8L, 23L, -7L, 13L)
str(b)

```

* creating vector 

```{r}

x <- c(1, 8, 23, -7, 13)
x <- 1:20
y <- seq(from = 3, to = 8, by = 0.2)


rep("Female", 10)

```

* Vectorization

```{r}

x <- c(3, 7, 6, 3, 5, 2)
x + 1
x*2
sqrt(x)

```


* logical vector


```{r}
x
z <- x > 4
z
y <- c("a", "b", "a", "d", "e")

t <- y == "A"
t
t <- y == "a"
t

```

* useful functions for vectors

```{r}

x
length(x)
sum(x)
max(x)
min(x)
sort(x)
order(x)
unique(x)

mean(x)
sd(x)


y
length(y)
unique(y)

```




* subset vector

```{r}
x
x[1]
x[3:5]
x[c(1,3:5)]

x[-2]
x[x > 4]


v <- 3
v[1]

```

### Matrices

* same type of data
* columns & rows


```{r}

x <- rbind(c(1:4), c(5:8))
x
dim(x)
dimnames(x)
attributes(x)

x <- cbind(c(1:4), c(5:8))
attributes(x)

x <- matrix(1:8, nrow = 2, ncol = 4, byrow = T)
x
x <- matrix(1:8, nrow = 2, ncol = 4, byrow = F)
x

```

### Data frame

* extension of matrix
* columns of different data types


```{r}
id <- 1:6
gender <- rep(c("F", "M"), each = 3)
age <- rep(8, 6)

students <- cbind(id, gender, age)
students
str(students)
summary(students)

students <- cbind.data.frame(id, gender, age)
students
str(students)
summary(students)

id <- 1:6
math <- c(9, 6, 7, 8, 10, 5)
english <- c(8, 5, 9, 8, 9, 7)

results <- cbind.data.frame(id, math, english)

df <- merge(students, results)
df

```
* subset data frame

```{r}

sub1 <- df[1:3, 1:2]
sub1

sub2 <- df[, c(1, 4:5)]
sub2

sub3 <- df$gender
sub3

sub4 <- subset(df, math > 7, select = c(id, english))
sub4

sub5 <- subset(df, english <= 7, select = -age)
sub5

```
* reorder the data frame

```{r}

df1 <- df[order(df$math),]
df1 
```
### List

* contain different type of data
* can contain data frame
```{r}
cls1 <- data.frame(id = 1:5, names = c("Lan", "Hung", "Tuan", "Mai", "Long"))
cls2 <- data.frame(id = 6:10, names = c("Thanh", "Son", "Nghia", "Hanh", "Thuy"))

ls <- list(cls1 = cls1, cls2 = cls2)
ls

```

* subset list

``` {r}
ls[1]
ls[[1]]

ls$cls1

```


## Import data

* Direct typing
* From clipboard

```{r}

# open excel file "C:\Users\nbngo\OneDrive\Work\[C]_ResearchMethods\quant_rm\Data\1 Raw\Dataset_environmental_sustainability.xlsx"

# env <- read.delim("clipboard")


```

* From csv 

``` {r}

env <- read.csv("Data/1 Raw/Dataset_environmental_sustainability.csv", sep = ",", header = T)


```

* From xlsx
``` {r}

# install.packages("readxl")
library(readxl)

env <- read_excel("Data/1 Raw/Dataset_environmental_sustainability.xlsx")

```

* From Rdata
``` {r}

load("Data/1 Raw/ntl_joined_avg.Rdata")


```

## Export data

* as Rdata
``` {r}

save(env, file = "Data/2 Processed/environment_survey.Rdata")


```

* to clipboard
```{r}
write.table(ntl_joined_avg, "clipboard", sep = "\t", row.names = F)
```

* to csv
```{r}
write.csv(ntl_joined_avg, file = "Data/2 Processed/ice_cover_lake.csv", row.names = F)
```


# Exploring and plotting data
## Data
* mtcars data in R


```{r}

data("mtcars")

head(mtcars)

```
* [explaining variables](https://rstudio-pubs-static.s3.amazonaws.com/61800_faea93548c6b49cc91cd0c5ef5059894.html)

* summary() gives overall information of the data

```{r}
summary(mtcars)
```

## Numerical/continuous/quantitative variables
```{r}

summary(mtcars$mpg)

mean(mtcars$mpg)
median(mtcars$mpg)

sd(mtcars$mpg)
var(mtcars$mpg)
quantile(mtcars$mpg, seq(0, 1, 0.2))

hist(mtcars$mpg)

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram()

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2)

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 2, fill = "white", color = "black") +
  theme_bw()

??theme_bw

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(aes(y = ..density..), binwidth = 2, fill = "white", color = "black") +
  geom_density(linewidth = 1) +
  theme_bw()


ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(aes(y = ..density..), binwidth = 2, fill = "white", color = "grey50") +
  geom_density(linewidth = 1) +
  geom_vline(xintercept = 19.20, color = "green", linewidth = 1) +
  geom_vline(xintercept = 20.09, color = "blue", linewidth = 1) +
  geom_vline(xintercept = c(15.43, 22.8), color = "purple", linewidth = 1) +
  theme_bw()

```
## Character/Factor (Nominal/Order/Categorical)

```{r}

table(mtcars$cyl)

prop.table(table(mtcars$cyl))

table(mtcars$am)
prop.table(table(mtcars$am))


ggplot(mtcars, aes(x = am)) +
  geom_bar(fill = "lightblue", color = "grey50") +
  theme_bw()

mtcars$am_f <- factor(mtcars$am)


ggplot(mtcars, aes(x = am_f)) +
  geom_bar(fill = "lightblue", color = "grey50") +
  theme_bw()

mtcars$am_f <- factor(mtcars$am, labels = c("auto", "manual"))


ggplot(mtcars, aes(x = am_f)) +
  geom_bar(fill = "lightblue", color = "grey50") +
  theme_bw()

ggplot(mtcars, aes(x = am_f)) +
  geom_bar(fill = "lightblue", color = "grey50") +
  theme_bw() +
  labs(x = "Transmission", y = "Count")

```

## Bivariate 
* continuous & discrete

```{r}


ggplot(mtcars, aes(x = am_f, y = mpg)) +
  geom_bar(stat = "summary", fun = "mean", fill = "lightblue", color = "grey50") +
  theme_bw() +
  labs(x = "Transmission", y = "Miles/gallon")

# install.packages("tidyverse")
library(tidyverse)


mpg_am <- mtcars %>%
  group_by(am_f) %>%
  summarise(mpg_mean = mean(mpg),
            mpg_sd = sd(mpg),
            mpg_min = min(mpg),
            mpg_max = max(mpg),
            mpg_q1 = quantile(mpg, 0.25),
            mpg_median = median(mpg),
            mpg_q3 = quantile(mpg, 0.75))

ggplot(mpg_am, aes(x = am_f, y = mpg_mean)) +
  geom_point(size = 3, shape = 18) +
  geom_errorbar(aes(ymin = mpg_mean - mpg_sd, ymax = mpg_mean + mpg_sd), width = 0.2) +
  theme_bw() +
  labs(x = "Transmission", y = "Miles/gallon")

ggplot(mtcars, aes(x = am_f, y = mpg)) +
  geom_boxplot() +
  theme_bw() +
  labs(x = "Transmission", y = "Miles/gallon")


```

* Continuous & continuous

```{r}

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  theme_bw() +
  labs(x = "Weight (1000 lbs)", y = "Miles/gallon")


ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  theme_bw() +
  labs(x = "Weight (1000 lbs)", y = "Miles/gallon")


cor(mtcars$wt, mtcars$mpg)
cor.test(mtcars$wt, mtcars$mpg)

```
* Discrete & discrete

```{r}

table(mtcars$cyl, mtcars$am_f)
prop.table(table(mtcars$cyl, mtcars$am_f), 1)
prop.table(table(mtcars$cyl, mtcars$am_f), 2)


ggplot(mtcars, aes(x = am_f, fill = cyl, group = cyl)) +
  geom_bar() +
  theme_bw() +
  labs(x = "Transmission", y = "Count", fill = "Number of cylinders")


mtcars$cyl_f <- factor(mtcars$cyl)

ggplot(mtcars, aes(x = am_f, fill = cyl_f, group = cyl)) +
  geom_bar() +
  theme_bw() +
  labs(x = "Transmission", y = "Count", fill = "Number of cylinders")

ggplot(mtcars, aes(x = am_f, fill = cyl_f, group = cyl)) +
  geom_bar(position = "fill") +
  theme_bw() +
  labs(x = "Transmission", y = "Proportion", fill = "Number of cylinders")



```

## Add dimension


```{r}

ggplot(mtcars, aes(x = wt, y = mpg, col = am_f)) +
  geom_point()  +
  theme_bw() +
  labs(x = "Weight (1000 lbs)", y = "Miles/gallon", col = "Transmission")

ggplot(mtcars, aes(x = wt, y = mpg, col = am_f)) +
  geom_point()  +
  geom_smooth(method = "lm", se = F) +
  theme_bw() +
  labs(x = "Weight (1000 lbs)", y = "Miles/gallon", col = "Transmission")

ggplot(mtcars, aes(x = wt, y = mpg, col = am_f)) +
  geom_point()   +
  facet_grid(.~cyl_f, labeller = labeller(.cols = label_both)) +
  theme_bw() +
  labs(x = "Weight (1000 lbs)", y = "Miles/gallon", col = "Transmission")



```


# Linear regression

## Compare means


```{r}

fit1 <- lm(mpg ~ am_f, mtcars)
summary(fit1)


```

## Simple linear regression


```{r}

fit2 <- lm(mpg ~ wt, mtcars)
summary(fit2)

```

## Multiple linear regression


```{r}

fit3 <- lm(mpg ~ wt + am_f, mtcars)
summary(fit3)

fit4 <- lm(mpg ~ wt + I(wt^2), mtcars)
summary(fit4)

AIC(fit2)
AIC(fit3)
AIC(fit4)

anova(fit3, fit2)
anova(fit4, fit2)



# plot(fit4)

```

# Factor analysis


```{r}

# install.packages("lavaan")
library(lavaan)


round(cor(env[,9:20], method = "spearman"),2)


att_bhv <- env[,9:20]


model1 <- '
  att =~ Q1 + Q2 + Q3 + Q4 + Q5 + Q6 + Q7
  bhv =~ Q8 + Q9 + Q10 + Q11 + Q12
'

cfa_result <- cfa(model1, data = att_bhv, std.lv=TRUE)
summary(cfa_result, fit.measures = T, standardized = T)



# install.packages("psych")
library(psych)
install.packages("GPArotation")
library(GPArotation)

att_bhv_sc <- scale(att_bhv)

fa <- fa(r = att_bhv_sc, nfactors = 2)
summary(fa)
fa$loadings



model2 <- '
  att =~ Q1 + Q3 + Q4 + Q5 + Q7
  bhv =~ Q8 + Q9 + Q11 + Q12
'

cfa_result <- cfa(model2, data = env, std.lv=TRUE)
summary(cfa_result, fit.measures = T, standardized = T)

model3 <- '
  att =~ Q1 + Q3 + Q5 
  bhv =~ Q8 + Q9 + Q10 + Q11 + Q12
'

cfa_result <- cfa(model3, data = env, std.lv=TRUE)
summary(cfa_result, fit.measures = T, standardized = T)

```

